\bf Range Analysis for the IonMonkey JavaScript engine

### 2012-05-02 project-writings/paper.tex
s/javascript/JavaScript/ig

\section{Introduction}
With the explosion of browser based applications that has occurred in
recent years, Javascript has become one of the most widely used
programming languages, and Javascript performance has become
critically important. Unfortunately, Javascript seems to be designed
to be un-performant.

One source of these woes are Javascript's dynamic semantics for
numbers: all numeric values are defined by the specification to be
IEEE 754 double precision floating point values (with bit-wise
operations being defined to round to the nearest 32 bit integer before
performing the operation). Since the common case is that numeric
values are integers and because floating point operations are
more expensive, all major Javascript implementations attempt to
represent numbers as integers when possible.

In IonMonkey, Mozilla's next-generation optimizing Javascript JIT
compiler, a hybrid type inference system (which uses both static analysis and
runtime information\cite{BHackettTI}) is used to generate
type-specialized code for the common case of integer arithmetic, and
type guards are inserted if the type inferencer is not certain. The
catch is that since double precision floats have a larger range than
32-bit integers, if the integer operations would overflow the result
must be promoted to being stored in a double. IonMonkey deals with
this by inserting overflow checks after arithmetic operations and
bailing out of the type specialized code if they trip. While this is
sometimes actually necessary, it is frequently the case that overflow
is not possible.

@@

\section{IonMonkey Overview}
The IonMonkey compiler is somewhat loosely based on LLVM, in that it is an
imperative SSA compiler. IonMonkey takes the bytecode used by the SpiderMonkey
Javascript interpreter and compiles it to native code. Optimistic type
specialization is used, and if type assumptions fail, then execution ``bails
out'' to the interpreter.

@@

While the ECMA specification for Javascript requires that all numbers behave as
double-precision floating point values\cite{ECMA-262}, for performance reasons JIT compilers
represent numbers as 32-bit integers when possible. However, since doubles can
represent larger values than integers, operations must be guarded against
overflow.  Our range analysis pass eliminates overflow checks when operations
can be proven not to overflow.

@@

This approach does not work for us for two reasons: first, IonMonkey
does not use Cytron et al.'s algorithm for computing SSA form; it
instead takes advantage of the structured nature of Javascript code to
go to SSA form without computing dominance frontiers \footnote{More
  precisely, it takes advantage in an ad-hoc and terrifying way of
  poorly documented structured invariants and ``source notes'' in the
  bytecode that the SpiderMonkey front-end emits.}.  Furthermore, we
decided that instead of putting the program into XSA form at the
beginning and leaving it there through the compilation phase, it would
be a better idea to introduce $\beta$ nodes solely for performing
range analysis and remove them afterwards. This is for two main
reasons: first, as a question of engineering effort, we did not want
to teach every pass of the compiler how to deal with $\beta$ nodes,
and the IonMonkey team does not want us to change the IL in a pervasive
way simply for range analysis; second, $\beta$ nodes are copies that
are not allowed to be propagated away, which may inhibit optimization
passes (like global value numbering).

@@

There are three benchmark suites which are generally used to measure Javascript
JIT compilers: the v8bench suite developed by Google, the SunSpider benchmark
suite developed by Apple, and the Kraken benchmark suite developed by Mozilla.

@@

\section{Future Work and Conclusions}
We implement an SSA-based range analysis pass which is capable of inferring
range information on integer operations. We use the results of this analysis to
perform some simple optimizations relevant to Javascript: we eliminate overflow
checks which can be proven redundant, and also eliminate a limited number of
array bounds checks.

>>>>

\section{Introduction}
With the explosion of browser based applications that has occurred in
recent years, JavaScript has become one of the most widely used
programming languages, and JavaScript performance has become
critically important. Unfortunately, JavaScript seems to be designed
to be un-performant.

One source of these woes are JavaScript's dynamic semantics for
numbers: all numeric values are defined by the specification to be
IEEE 754 double precision floating point values (with bit-wise
operations being defined to round to the nearest 32 bit integer before
performing the operation). Since the common case is that numeric
values are integers and because floating point operations are
more expensive, all major JavaScript implementations attempt to
represent numbers as integers when possible.

In IonMonkey, Mozilla's next-generation optimizing JavaScript JIT
compiler, a hybrid type inference system (which uses both static analysis and
runtime information\cite{BHackettTI}) is used to generate
type-specialized code for the common case of integer arithmetic, and
type guards are inserted if the type inferencer is not certain. The
catch is that since double precision floats have a larger range than
32-bit integers, if the integer operations would overflow the result
must be promoted to being stored in a double. IonMonkey deals with
this by inserting overflow checks after arithmetic operations and
bailing out of the type specialized code if they trip. While this is
sometimes actually necessary, it is frequently the case that overflow
is not possible.

@@

\section{IonMonkey Overview}
The IonMonkey compiler is somewhat loosely based on LLVM, in that it is an
imperative SSA compiler. IonMonkey takes the bytecode used by the SpiderMonkey
JavaScript interpreter and compiles it to native code. Optimistic type
specialization is used, and if type assumptions fail, then execution ``bails
out'' to the interpreter.

@@

While the ECMA specification for JavaScript requires that all numbers behave as
double-precision floating point values\cite{ECMA-262}, for performance reasons JIT compilers
represent numbers as 32-bit integers when possible. However, since doubles can
represent larger values than integers, operations must be guarded against
overflow.  Our range analysis pass eliminates overflow checks when operations
can be proven not to overflow.

@@

This approach does not work for us for two reasons: first, IonMonkey
does not use Cytron et al.'s algorithm for computing SSA form; it
instead takes advantage of the structured nature of JavaScript code to
go to SSA form without computing dominance frontiers \footnote{More
  precisely, it takes advantage in an ad-hoc and terrifying way of
  poorly documented structured invariants and ``source notes'' in the
  bytecode that the SpiderMonkey front-end emits.}.  Furthermore, we
decided that instead of putting the program into XSA form at the
beginning and leaving it there through the compilation phase, it would
be a better idea to introduce $\beta$ nodes solely for performing
range analysis and remove them afterwards. This is for two main
reasons: first, as a question of engineering effort, we did not want
to teach every pass of the compiler how to deal with $\beta$ nodes,
and the IonMonkey team does not want us to change the IL in a pervasive
way simply for range analysis; second, $\beta$ nodes are copies that
are not allowed to be propagated away, which may inhibit optimization
passes (like global value numbering).

@@

There are three benchmark suites which are generally used to measure JavaScript
JIT compilers: the v8bench suite developed by Google, the SunSpider benchmark
suite developed by Apple, and the Kraken benchmark suite developed by Mozilla.

@@

\section{Future Work and Conclusions}
We implement an SSA-based range analysis pass which is capable of inferring
range information on integer operations. We use the results of this analysis to
perform some simple optimizations relevant to JavaScript: we eliminate overflow
checks which can be proven redundant, and also eliminate a limited number of
array bounds checks.
